---
title: 'Introduction to Logistic Regression with R'
author: 'Wilbur Ouma'
date: 'March 22, 2024'
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

## Overview

In this workshop, we will introduce the logistic regression model as
special case of a linear regression model (a generalized linear model)
where the **response variable is categorical**.

We will discuss responses in terms of **probabilities and odds**.

## Objectives

Upon successful completion of this workshop, you should be able to:

-   Understand the basics of the logistic regression model

-   Fit a logistic model to data to model a binary response variable as
    a function of a set of explanatory variables.

-   Interpret coefficients of a logistic regression model

    -   Understand odds and odds ratios

-   Test and asses model accuracy with drop-in-deviance likelihood ratio
    tests

## Model definition

Logistic regression models a binary response variable as a function of
explanatory/predictor variable(s).

For instance, given the categorical response variable $Y$ for *Disease
State* taking on the values of *Disease* and *No disease*:

$\mathrm{Y} = \begin{cases} 1 & \text{Disease} \\ 0 & \text{No disease} \end{cases}$

we no longer model $Y = \mu(Y|X)$

Rather we model ***probability*** that $Y$ belongs to a particular
category/class: $Pr(Y = 1| X)$, for instance the probability of
developing the disease.

For logistic regression --- with one explanatory variable --- we model
$Pr(Y = 1| X)$ using a function that gives outputs between 0 and 1.

The **logistic function** below describes a population proportion or
probability as a function of one predictor variable:

$Pr(Y = 1| X) = \frac {e^{\beta_0 + \beta_1X_1}} {1 + e^{\beta_0 + \beta_1X_1}}$

A logistic function is similar to a common S-shaped (sigmoid) curve:

$S(x)=\frac{e^x}{1+e^x}$

![Sigmoid curve. Image source:
Wikipedia](images/paste-E3D173A3.png){width="430"}

### The logit (log odds) function

Manipulating the logistic function encountered above yields:

$\frac {Pr(Y = 1| X)}{1-Pr(Y = 1| X)} = e^{\beta_0 + \beta_1X}$

The quantity $\frac {Pr(Y = 1| X)}{1-Pr(Y = 1| X)}$ is the **odds**
($\omega$)

Taking natural logs yields a **log odds (logit)** function:

$ln(\frac {Pr(Y = 1| X)}{1-Pr(Y = 1| X)}) = ln(\omega) = \beta_0 + \beta_1X$

We can now interpret the coefficient $\beta_1$ of the logit function as
follows:

**Increasing** $X$ **by one unit changes the log odds by**
$\beta_1$**.**

**Equivalently, it multiplies the odds by** $e^{\beta_1}$

The same formalism extends to multiple *p* explanatory variables
(continuous and categorical):

$ln(\omega) = \beta_0 + \beta_1X_1 + \beta_2X_2 + ...+\beta_pX_p$

Exponentiating the logit yields the odds:

$\omega = e^{\beta_0 + \beta_1X_1 + \beta_2X_2 + ...+\beta_pX_p}$

We can then compare the **odds** between different values of a
particular predictor $X$, for fixed values of other predictors.

For instance, the odds at $X_1=A$ relative to the odds at $X_1=B$ ---
for fixed values of other $Xs$ --- is:

$\phi = \frac {\omega_A}{\omega_B} = e^{[\beta_1(A-B)]}$

Unlike the case in simple and multiple linear regression, coefficients
in logistic regression are estimated by the maximum likelihood method.

### Odds and Odds Ratio (OR): A Review

Consider the following data set showing the number of individuals that
developed coronary artery disease (CAD).

| Gender | Disease   | No Disease |
|--------|-----------|------------|
| Male   | 217 (57%) | 162 (43%)  |
| Female | 105 (44%) | 136 (56%)  |

Gender is a potential (binary) predictor of CAD, and the response
variable is also binary (Disease/No Disease).

To determine whether the two variables are independent, we can perform a
Chi-square test of independence.

```{r}
#create a table of data
cad <- as.table(rbind(c(217, 162), c(105, 136)))
dimnames(cad) <- list(Gender = c("M", "F"),
                    CAD = c("Disease","No Disease"))
cad
```

```{r}
#Perform a Chi-sqr test
cad_Xsq <- chisq.test(cad)
cad_Xsq

```

There's evidence for a significant association between gender and the
likelihood of developing CAD (p-value = 0.0012).

Since we have only one qualitative predictor (no quantitative
predictor), the Chi-square test is sufficient, no need for logistic
regression. Otherwise a logistic regression with multiple predictors
would be used.

Notice how the above test is essentially comparing proportions of two
groups (Male and Female), a comparison that is analogous to comparing
two means such as in t-tests or SLR model.

We can quantify the significant Gender effect with Odds ratios.

| Gender | Probability of Disease | Odds ($\hat\omega$)     |
|--------|------------------------|-------------------------|
| Male   | 0.573                  | 0.573/(1-0.573) = 1.341 |
| Female | 0.436                  | 0.436/(1-0.436) = 0.773 |

$OR = \hat\phi = \frac {\hat\omega_M}{\hat\omega_F} = 1.341\div 0.773 = 1.74$

The odds of men developing CAD are estimated to be 1.74 times as large
as the odds of women.

Equivalently, the odds of men developing CAD are
($100 \times (\hat\phi - 1) = 100 \times (1.74 -1)$) 74% greater than
the odds of women developing CAD.

In the following case study, we incorporate multiple predictors to model
their effect on log odds of a binary response variable.

## Case Study 1: Survival in the Donner Party

In 1846 the Donner and Reed families left Springfield, Illinois, for
California by covered wagon. In July, the Donner Party, as it became
known, reached Fort Bridger, Wyoming. There, its leaders decided to
attempt a new and untested route to the Sacramento Valley. Having
reached its full size of 87 people and 20 wagons, the party was delayed
by a difficult crossing of the Wasatch Range and again in the crossing
of the desert west of the Great Salt Lake. The group became stranded in
the eastern Sierra Nevada mountains when the region was hit by heavy
snows in late October. By the time the last survivor was rescued on
April 21, 1847, 40 of the 87 members had died from famine and exposure
to extreme cold.

If you like history podcast, listen to **"The Donner Party"** story in
the **Legends of the Old West** Podcast.

In 1990, an anthropologist wanted to study the theory that females are
better able to withstand harsh conditions than are males. The data frame
containing the ages and sexes of the adult (over 15 years) survivors and
nonsurvivors of the Donner party is sound in the `Sleuth3` R package
(case2001).

**Research question:** **After accounting for the effects of age, did
women have greater survival odds than men?**

To model the effects of Age and Sex on survival, we fit a log odds
(logit) function:

$ln(\omega) = \beta_0 + \beta_1Age + \beta_2Sex$,

where
$\omega = \frac {Pr(Status=survived | Age, Sex)}{1-Pr(Status=survived | Age, Sex)}$

We interpret $\beta_1$ as the change in **log odds** of survival
associated with one year increase in age.

Equivalently, an increase of one year in age multiplies survival
**odds** by $e^{\beta_1}$.

To answer our research question, we compare odds of survival for two
groups (males vs females) --- *after accounting for odds of survival
associated with age* --- and we are therefore interested in **odds
ratios** ($\phi$).

Notice that Sex is a categorical predictor variable. Therefore, we
create an indicator/dummy variable such that:

$\mathrm{sexFemale} = \begin{cases} 1 & \text{Female} \\ 0 & \text{Male} \end{cases}$

The logit function becomes:

$ln(\omega) = \beta_0 + \beta_1Age + \beta_2SexFemale = \mathrm{} \begin{cases} \beta_0 + \beta_1Age + \beta_2 = ln(\omega_F) & \text{Female log odds} \\ \beta_0 + \beta_1Age =ln(\omega_M) & \text{Male log odds} \end{cases}$

The log odds ratio (log odds of female Vs log odds of males) becomes:

$ln(\phi) = ln(\frac {\omega_F}{\omega_M}) = ln(\omega_F) - ln(\omega_M) = \beta_2$

The odds ratio becomes:

$\phi = e^{\beta_2}$

We will obtain the (log) odds ratio from the logit fit in `R`.

We first load required packages.

```{r, warning=FALSE}
#load required packages
library(tidyverse)
library(Sleuth3)
library(ISLR2)
```

And load the data set

```{r}
donner_data <- case2001
print.data.frame(head(donner_data))
```

The data set has 45 observations with two predictor variables (Age, Sex)
and a binary categorical response variable (Status).

We first make "Male" the reference level of the Sex predictor variable.

```{r}
donner_data <- donner_data %>%
  mutate(Sex = relevel(Sex, ref = "Male"))

str(donner_data)
```

We then fit an additive logit model for survival odds, using **a
Generalized Linear Model (GLM)** function, and estimate model parameters
using the **method of maximum likelihood**.

```{r}
fit_donner<- glm(Status ~ Age + Sex, data = donner_data, 
                               family=binomial)
summary(fit_donner)
```

#### **Effect of age on survival**

The **log odds** of survival associated with one year increase in age
are $\hat\beta_1 = -0.0782$.

The **odds of s**urvival associated with one year increase in age are
$e^{\hat\beta_1}=e^{-0.0782}=0.925$

Equivalently, one year increase in age multiplies the odds of survival
by 0.925, i.e the survival odds change by
$100\times(e^{\hat\beta_1}-1) = 100\times(0.925-1) = -7\%$, implying **a
reduction of 7%** in the odds of surviving.

Recall how this interpretation of coefficients is similar to the
interpretation encountered when the response variable $Y$ is
log-transformed in the case of a simple/multiple linear regression
model.

**95% Confidence Intervals**

We can obtain the 95% confidence intervals of the estimated coefficient
for age:

```{r}
#95% CI of the log odds
confint(fit_donner, 2)
#95% CI of the odds
exp(confint(fit_donner, 2))
```

#### **Effect of Sex on survival: did women have greater survival odds than men?**

We compare the odds of survival for women and men, and are therefore
interested in the (log) odds ratio associated with the coefficient for
SexFemale ($\beta_2$):

$\phi = e^{\beta_2}$

The estimated **log odds ratio** (from the logit function above) is
1.5972 (p-value = 0.0345).

The **log odds** of survival for females are estimated to be 1.5972
times the log odds of survival for men.

The **odds** of survival for females are estimated to be 5-times
($e^{1.5972}=4.9392$) the odds of survival for males of similar age (95%
CI: 1.2 times to 25.2 times, see below).

```{r}
exp(confint(fit_donner,3))
```

### Learning Check!

What are the odds of survival for a 55-year-old relative to the odds of
survival for a 30-year-old? What is the associated 95% CI?

Hint: Recall ---

"We can then compare the odds between different values of a particular
predictor $X$, for fixed values of other predictors. For instance, the
odds at $X_1=A$ relative to the odds at $X_1=B$ --- for fixed values of
other $Xs$ --- is:"

$\phi = \frac {\omega_A}{\omega_B} = e^{[\beta_1(A-B)]}$

### Drop-in-deviance test: A Likelihood Ratio Test

Sometimes when performing data analysis involving multiple predictors,
we want to judge the adequacy of a reduced model relative to a full
model, i.e. we want to determine whether inclusion of additional
predictors (and their parameters/coefficients) is warranted.

Given these two logit models:

Full model ($H_A$): $ln(\omega) = \beta_0 + \beta_1X_1 + \beta_2X_2$

Reduced model ($H_0$): $ln(\omega) = \beta_0 + \beta_1X_1$

where the reduced model is a special case (nested) of the full model, we
can determine whether inclusion of the term $\beta_2X_2$ is warranted in
explaining the log odds.

A **Likelihood ratio test** compares the maximized likelihood ($LMAX$)
of parameters in the full versus reduced model, such that:

$LRT=2\times ln(LMAX_{Full} - 2\times ln(LMAX_{Reduced})$

When the reduced model is correct (under $H_0$):

$LRT \sim \chi^2$, with d.f equal to \# of parameters in full minus
reduced model.

We can obtain approximate p-value as the proportion of the $\chi^2$
distribution that is greater than $LRT$.

Some statistical packages provide a quantity called **deviance** rather
than the maximized likelihood, wherein

$deviance=constant-2\times ln(LMAX)$, and

$LRT = deviance_{reduced}-deviance_{full}$

## Case Study 2: Birdkeeping and Lung Cancer --- A Retrospective Observational Study

A 1972--1981 health survey in The Hague, Netherlands, discovered an
association between keeping pet birds and increased risk of lung cancer.

To investigate birdkeeping as a risk factor, researchers conducted **a
case--control study** of patients in 1985 at four hospitals in The Hague
(population 450,000).

They identified **49 cases of lung cancer** among patients who were
registered with a general practice, who were age 65 or younger, and who
had resided in the city since 1965.

They also selected **98 controls** from a population of residents having
the same general age structure. (Data based on *P. A. Holst, D.
Kromhout, and R. Brand, "For Debate: Pet Birds as an Independent Risk
Factor for Lung Cancer," British Medical Journal 297 (1988): 13--21*.)

Data was gathered on the following variables:

1.  ð¿ð¶: Lung cancer (ð¿ð¶=1 for lung cancer patient, 0 for control)

2.  ð¹ð‘€: Sex (ð¹=1, ð‘€=0)

3.  ð´ðº: Age, in years

4.  ð‘†ð‘†: Socioeconomic status (ð»ð‘–ð‘”â„Ž=1, ð¿ð‘œð‘¤=0)

5.  ð‘Œð‘…: Years of smoking prior to diagnosis or examination

6.  ðµð¾: Indicator of birdkeeping

7.  CD: Average rate of smoking (in cigarettes per day)

Data is contained in the `Sleuth3` R package (case2002).

**Research question: after age, socioeconomic status, and smoking have
been controlled for, is [*an additional risk*]{.underline} of developing
lung cancer associated with birdkeeping?**

We will answer this question by performing a drop-in-deviance test to
determine whether the term for birdkeeping is significant.

$H_0: ln(\omega_{LC}) = \beta_0 + \beta_1FM + \beta_2SS + \beta_3AG + \beta_4YR + \beta_5CD$

$H_A: ln(\omega_{LC}) = \beta_0 + \beta_1FM + \beta_2SS + \beta_3AG + \beta_4YR + \beta_5CD + \beta_6BK$

We load the data, and reorder the levels so that the model is for log
odds of cancer, and for BK categorical variable make "NoBird" reference
level.

```{r}
birdkeeping<-case2002
birdkeeping$LC <- factor(birdkeeping$LC, levels=c("NoCancer","LungCancer"))

birdkeeping <- birdkeeping %>%
  mutate(BK = relevel(BK, ref = "NoBird"))

print.data.frame(head(birdkeeping))
str(birdkeeping)
```

Next, we fit a full logit model to the data:

```{r}
fit_birdkeeping <- glm(LC ~ FM + SS + AG + YR + CD + BK,
                         data = birdkeeping, family=binomial)

summary(fit_birdkeeping)
```

It appears that there's an effect of Years of Smoking and of Bird
Keeping after accounting for other variables; no obvious effects of
other variables.

#### A Strategy for Variable Selection: backward elimination

We can start building a logistic regression model for $H_0$ using
backward elimination (**withholding BK**), until we remain with
significant predictors.

We sequentially remove variables based on p-value of their coefficients.

```{r}
myGlm1 <- glm(LC ~ FM + SS + AG + YR + CD, data = birdkeeping, family=binomial)
summary(myGlm1)
```

```{r}
myGlm2 <- update(myGlm1, ~ . - SS)        
summary(myGlm2)
```

```{r}
myGlm3 <- update(myGlm2, ~ . - CD)   
summary(myGlm3)
```

```{r}
myGlm4 <- update(myGlm3, ~ . - FM)
summary(myGlm4) # Everything left has a small p-value (retain the intercept)
```

We now re-fit our full model $H_A$ **to include BK**.

```{r}
fit_birdkeeping <- update(myGlm4, ~ . + BK)
summary(fit_birdkeeping)
```

**Research question:** after age, socioeconomic status, and smoking have
been controlled for, is **an additional risk** of developing lung cancer
associated with birdkeeping?

And test for the birdkeeping effect using the drop-in-deviance LRT:

$H_0: ln(\omega_{LC}) = \beta_0 + \beta_1AG + \beta_2YR$

$H_A: ln(\omega_{LC}) = \beta_0 + \beta_1AG + \beta_2YR + \beta_3BK$

$LRT = deviance_{H_0}-deviance_{H_A}$

```{r}
anova(myGlm4, fit_birdkeeping) # Drop-in-deviance = 12.612 on 1 d.f.

```

And we calculate **a 1-sided p-value** (question is on additional risk,
rather than differences in risk)

```{r}
(1 - pchisq(12.612,1))/2 #p-value: 0.0001916391
```

**Interpretation:**

There is evidence for additional risk of developing lung cancer
associated with birdkeeping, after age and years of smoking have been
accounted for (one-sided p-value = 0.000192).

What are the odds of lung cancer for people who kept birds compared to
those who didn't, after accounting for other variables?

```{r}
beta <- fit_birdkeeping$coef  # Extract estimated coefficients
beta
exp(beta[4])   # 3.961248                
exp(confint(fit_birdkeeping,4))   # 1.836764 8.900840  
```

The odds of lung cancer for people who kept birds were estimated to be
**4 times** the odds of lung cancer for people of similar age, sex, \#
smoking history, and socio-economic status who didn't keep birds.

### Conclusion

-   When we encounter a data set with a binary response variable and a
    set of predictor variables, we can model the odds of the response as
    function of the predictors using the logit function.

    -   When only one categorical predictor is used, then a Chi-square
        test for independence would suffice.

-   The coefficients of the predictors of the logit function denote the
    log odds for the response with each unit increase in the value of
    continuous predictor; or the log odds ratio in case of a categorical
    predictor.

-   Odds are obtained by exponentiating the estimated coefficients of
    the logit function.
